Fully Extrinsic Type Inference (DRAFT 2025-06-17)

### 2025-07-20 doc/extrinsic_type_inference_paper/paper.tex
add notes on variable terminology
update to make locale the term for the triple of skolems, constraints, and a type

@@

\begin{definition}
  \label{def:sequence}
  \small
  \nopad
  \[\begin{array}[t]{rr@{}l}
    \com{type var seq} &
    \nonterm{\Theta} & \is \epsilon \sep \Theta\ \alpha 
    \\
    \com{subtyping seq} &
    \nonterm{\Delta} & \is \epsilon \sep \Delta \  \tau \J{<:} \tau
    \\
    \com{typing seq} &
    \nonterm{\Gamma} & \is \epsilon \sep \Gamma\ x \hastype \tau
    \\
    \com{schema seq} &
    \nonterm{\Pi} & \is \epsilon \sep \Pi\ (\Theta, \Delta, \tau) 
    \\
    \com{subtractable type seq} &
    \nonterm{\Xi} & \is \epsilon \sep \Xi\ \eta 
    \\
    \com{negatable subtyping seq} &
    \nonterm{\Omega} & \is \epsilon \sep 
      \Omega\ \tau \J{<:} \eta \sep 
      \Omega\ \alpha \J{<:} \tau
    \\
    \com{composite seq} &
    \nonterm{C} & \is \epsilon \sep C\ \iota 
    \\
    \com{item} &
    \nonterm{\iota} & \is 
    \alpha
    \sep 
    \tau \J{<:} \tau
    \sep
    x \hastype \tau
    \sep
    (\Theta, \Delta, \tau)
    \sep
    \eta
    \sep
    \tau \J{<:} \eta

@@

For a function \ms{f},  
it delegates the work to two helpers. 
First, it leverages function typing consolidation 
\ms{\Theta, \Delta, \Gamma \entails f \liftfun \Pi \sim \Xi}
to infer as many types as possible
and consolidates the results into a sequence of schemas \ms{\Pi}.
It must consolidate all of the inferred types in order to
ensure that the inferred type of each 
path's pattern does not subtype the type of subsequent paths' patterns. 
The sequence of subtractable types \ms{\Xi} can be ignored, as it is only
exposed for the sake of being used recursively.
Next, it leverage positive packing 
\ms{\textbf{pack}^+(... \entails \Pi) = \tau}
to rewrite each schema into a type,
such that the skolems and plastics are converted to
existential and universal bindings,
and then combine all the types by compounding intersection.
Packing the path types together ensures that certain
control flows of application can be verified even when  
the best approximation of the type of the argument
is less precise than the pattern of any one path.

For projection \ms{e\J{.}l}, 
it delegates to proof subtyping 
\ms{... \entails \tau \subtypes \J{<}l\J{>}\alpha \given ...}
to check that there is an entry with the label \ms{l}
in the source expression's type \ms{\tau}.
With subtyping, it learns a lower bound on the type of the entry's content \ms{\alpha}.
For application \ms{e_f\J{(}e_a\J{)}}, 
it delates to proof subtyping 
\ms{... \entails \tau_f \subtypes \tau_a \J{->} \alpha \given ...}
to check that that the function \ms{e_f} can actually
map the argument \ms{e_a} to a result. With subtyping, 
it learns a lower bound on the type of the result \ms{\alpha}.
For an annotation \ms{\J{def}\ x \J{:} \tau_a\ \J{=}\ e\ \J{in}\ e'}, 
it first delegates to a few helpers. 
First it leverages typing consolidation \ms{\Theta, \Delta, \Gamma \entails e \hastype \Pi}
to infer as many possible schemas for the source expression \ms{e}.
Then it packs all the schemas together and ensures that the packed type
subtypes the type annotation.
It assigns the annotated type to the defined variable in the contextual typing
as it proceeds to infer the type of the continuation \ms{e'}, thereby
artificially restricting the permissible control flow.

For a loop \ms{\J{loop(}e\J{)}}, 
it eagerly factors the paths of the recursive function 
to construct a single path type 
with an recursive type as its antecedent and an recursive type as its consequent. 
First, it infers the type of the fixed point combinators argument \ms{e},
and it leverages loop subtyping consolidation 
\ms{\Theta', \Delta' \entails \tau \subtypes \alpha_\nu \rightarrow \Pi^\nu}
to ensure the argument has a  
form compatible with looping and 
to collect as many type schemas of its paths as possible.  
It leverage schema sequence duality 
\ms{
  \alpha_\nu \downarrow \Pi_\nu
  \fallingdotseq 
  \alpha_\mu \uparrow \Pi_\mu
}, which rewrites the implicitly co-recursive path schemas into
implicitly recursive pair schemas. 
Then, it leverages negative packing \ms{\textbf{pack}^-(... \entails \Pi_\mu) = \tau_\mu},
and constructs an explicitly recursive type by wrapping it in a least fixed point
\ms{\J{LFP[}\alpha_\mu\J{]}\tau_\mu}.
It then leverages factorization \ms{
  \Vdash \J{LFP[} \alpha_\mu \J{]}\tau_\mu \subtypes \J{<left>}\ \tau_l
} to factor out the left part of the recursive pair, and does the same for the right part. 
Finally from the left and right parts, it forms a single path type
with a least fixed point for both the antecedent and the consequent.

@@

\begin{requirement}
  \label{req:schema_sequence_duality_correspondence}
  \emph{schema sequence duality correspondence}
  \footnotesize
  \nopad
  \begin{mathpar}
    \inferrule {
      \alpha_\nu \downarrow \Pi_\nu \fallingdotseq \alpha_\mu \uparrow \Pi_\mu
      \\
      \delta \satisfies \Delta
      \\
      \alpha_\nu \notin \textbf{ftv}(\Delta)
      \\
      \alpha_\mu \notin \textbf{ftv}(\Delta)
    } {
      (
      \forall \Theta^\dagger, \Delta^\dagger, \tau^\dagger \qua
      (\Theta^\dagger,\Delta^\dagger, \tau^\dagger) \in \Pi_\nu \implies
      \exists \delta^\dagger \qua \textbf{dom}(\delta^\dagger) \subseteq \Theta^\dagger \up
      \\\\
      (
      \forall \delta' \qua
      \textbf{dom}(\delta') \cap \textbf{ftv}(\Delta) = \epsilon \implies
      % \delta\ \alpha_\nu \slash \J{BOT} \oplus \delta' \oplus \delta^\dagger \satisfies \Delta^\dagger \implies
      % \delta\ \alpha_\nu \slash \J{BOT} \oplus \delta' \oplus \delta^\dagger \satisfies e_f \hastype \tau^\dagger
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies \Delta \cup \Delta^\dagger \implies
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies e_f \hastype \tau^\dagger
      )
      )
      \\\\
      \iff
      \\\\
      (\forall e_a, e_r \qua
      (
      \exists \Theta^\dagger, \Delta^\dagger, \tau^\dagger \qua
      (\Theta^\dagger,\Delta^\dagger, \tau^\dagger) \in \Pi_\mu \up
      (\forall \delta^\dagger \qua \textbf{dom}(\delta^\dagger) \subseteq \Theta^\dagger \implies
      \\\\
      \exists \delta' \qua
      \textbf{dom}(\delta') \cap \textbf{ftv}(\Delta) = \epsilon \up
      % \delta\ \alpha_\mu \slash \J{TOP} \oplus \delta' \oplus \delta^\dagger \satisfies \Delta^\dagger \up
      % \delta\ \alpha_\mu \slash \J{TOP} \oplus \delta' \oplus \delta^\dagger \satisfies (e_a \J{,} e_r) \hastype \tau^\dagger 
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies \Delta \cup \Delta^\dagger \up
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies (e_a \J{,} e_r) \hastype \tau^\dagger 
      )
      )
      \implies
      e_r \cong e_f\J{(}e_a\J{)}
      )
    }
  \end{mathpar}
\end{requirement}

\noindent
The requirement of schema sequence duality has an ungodly amount of symbolic clutter,
but the essence of its meaning is actually quite simple and natural.
It says that a function inhabits all schemas in some 
some collection if and only if for any input/ouput pair, if the pair inhabits some
schema from the dual collection, then the function applied to the input
gives the output. 
The requirement is essential to ensure that rewriting inferred schemas
to construct least fixed point types does not violate soundness.

>>>>

\TODO{Other options are "unknown", "hole", "logic variable", "placeholder", "localetic variable"}

\TODO{replace "plastic" with "placeholder"}

@@

\begin{definition}
  \label{def:sequence}
  \small
  \nopad
  \[\begin{array}[t]{rr@{}l}
    \com{type var seq} &
    \nonterm{\Theta} & \is \epsilon \sep \Theta\ \alpha 
    \\
    \com{subtyping seq} &
    \nonterm{\Delta} & \is \epsilon \sep \Delta \  \tau \J{<:} \tau
    \\
    \com{typing seq} &
    \nonterm{\Gamma} & \is \epsilon \sep \Gamma\ x \hastype \tau
    \\
    \com{locale seq} &
    \nonterm{\Pi} & \is \epsilon \sep \Pi\ (\Theta, \Delta, \tau) 
    \\
    \com{subtractable type seq} &
    \nonterm{\Xi} & \is \epsilon \sep \Xi\ \eta 
    \\
    \com{negatable subtyping seq} &
    \nonterm{\Omega} & \is \epsilon \sep 
      \Omega\ \tau \J{<:} \eta \sep 
      \Omega\ \alpha \J{<:} \tau
    \\
    \com{composite seq} &
    \nonterm{C} & \is \epsilon \sep C\ \iota 
    \\
    \com{item} &
    \nonterm{\iota} & \is 
    \alpha
    \sep 
    \tau \J{<:} \tau
    \sep
    x \hastype \tau
    \sep
    (\Theta, \Delta, \tau)
    \sep
    \eta
    \sep
    \tau \J{<:} \eta

@@

For a function \ms{f},  
it delegates the work to two helpers. 
First, it leverages function typing consolidation 
\ms{\Theta, \Delta, \Gamma \entails f \liftfun \Pi \sim \Xi}
to infer as many types as possible
and consolidates the results into a sequence of locales \ms{\Pi}.
It must consolidate all of the inferred types in order to
ensure that the inferred type of each 
path's pattern does not subtype the type of subsequent paths' patterns. 
The sequence of subtractable types \ms{\Xi} can be ignored, as it is only
exposed for the sake of being used recursively.
Next, it leverage positive packing 
\ms{\textbf{pack}^+(... \entails \Pi) = \tau}
to rewrite each locale into a type,
such that the skolems and plastics are converted to
existential and universal bindings,
and then combine all the types by compounding intersection.
Packing the path types together ensures that certain
control flows of application can be verified even when  
the best approximation of the type of the argument
is less precise than the pattern of any one path.

For projection \ms{e\J{.}l}, 
it delegates to proof subtyping 
\ms{... \entails \tau \subtypes \J{<}l\J{>}\alpha \given ...}
to check that there is an entry with the label \ms{l}
in the source expression's type \ms{\tau}.
With subtyping, it learns a lower bound on the type of the entry's content \ms{\alpha}.
For application \ms{e_f\J{(}e_a\J{)}}, 
it delates to proof subtyping 
\ms{... \entails \tau_f \subtypes \tau_a \J{->} \alpha \given ...}
to check that that the function \ms{e_f} can actually
map the argument \ms{e_a} to a result. With subtyping, 
it learns a lower bound on the type of the result \ms{\alpha}.
For an annotation \ms{\J{def}\ x \J{:} \tau_a\ \J{=}\ e\ \J{in}\ e'}, 
it first delegates to a few helpers. 
First it leverages typing consolidation \ms{\Theta, \Delta, \Gamma \entails e \hastype \Pi}
to infer as many possible locales for the source expression \ms{e}.
Then it packs all the locales together and ensures that the packed type
subtypes the type annotation.
It assigns the annotated type to the defined variable in the contextual typing
as it proceeds to infer the type of the continuation \ms{e'}, thereby
artificially restricting the permissible control flow.

For a loop \ms{\J{loop(}e\J{)}}, 
it eagerly factors the paths of the recursive function 
to construct a single path type 
with an recursive type as its antecedent and an recursive type as its consequent. 
First, it infers the type of the fixed point combinators argument \ms{e},
and it leverages loop subtyping consolidation 
\ms{\Theta', \Delta' \entails \tau \subtypes \alpha_\nu \rightarrow \Pi^\nu}
to ensure the argument has a  
form compatible with looping and 
to collect as many type locales of its paths as possible.  
It leverage locale sequence duality 
\ms{
  \alpha_\nu \downarrow \Pi_\nu
  \fallingdotseq 
  \alpha_\mu \uparrow \Pi_\mu
}, which rewrites the implicitly co-recursive path locales into
implicitly recursive pair locales. 
Then, it leverages negative packing \ms{\textbf{pack}^-(... \entails \Pi_\mu) = \tau_\mu},
and constructs an explicitly recursive type by wrapping it in a least fixed point
\ms{\J{LFP[}\alpha_\mu\J{]}\tau_\mu}.
It then leverages factorization \ms{
  \Vdash \J{LFP[} \alpha_\mu \J{]}\tau_\mu \subtypes \J{<left>}\ \tau_l
} to factor out the left part of the recursive pair, and does the same for the right part. 
Finally from the left and right parts, it forms a single path type
with a least fixed point for both the antecedent and the consequent.

@@

\begin{requirement}
  \label{req:locale_sequence_duality_correspondence}
  \emph{locale sequence duality correspondence}
  \footnotesize
  \nopad
  \begin{mathpar}
    \inferrule {
      \alpha_\nu \downarrow \Pi_\nu \fallingdotseq \alpha_\mu \uparrow \Pi_\mu
      \\
      \delta \satisfies \Delta
      \\
      \alpha_\nu \notin \textbf{ftv}(\Delta)
      \\
      \alpha_\mu \notin \textbf{ftv}(\Delta)
    } {
      (
      \forall \Theta^\dagger, \Delta^\dagger, \tau^\dagger \qua
      (\Theta^\dagger,\Delta^\dagger, \tau^\dagger) \in \Pi_\nu \implies
      \exists \delta^\dagger \qua \textbf{dom}(\delta^\dagger) \subseteq \Theta^\dagger \up
      \\\\
      (
      \forall \delta' \qua
      \textbf{dom}(\delta') \cap \textbf{ftv}(\Delta) = \epsilon \implies
      % \delta\ \alpha_\nu \slash \J{BOT} \oplus \delta' \oplus \delta^\dagger \satisfies \Delta^\dagger \implies
      % \delta\ \alpha_\nu \slash \J{BOT} \oplus \delta' \oplus \delta^\dagger \satisfies e_f \hastype \tau^\dagger
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies \Delta \cup \Delta^\dagger \implies
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies e_f \hastype \tau^\dagger
      )
      )
      \\\\
      \iff
      \\\\
      (\forall e_a, e_r \qua
      (
      \exists \Theta^\dagger, \Delta^\dagger, \tau^\dagger \qua
      (\Theta^\dagger,\Delta^\dagger, \tau^\dagger) \in \Pi_\mu \up
      (\forall \delta^\dagger \qua \textbf{dom}(\delta^\dagger) \subseteq \Theta^\dagger \implies
      \\\\
      \exists \delta' \qua
      \textbf{dom}(\delta') \cap \textbf{ftv}(\Delta) = \epsilon \up
      % \delta\ \alpha_\mu \slash \J{TOP} \oplus \delta' \oplus \delta^\dagger \satisfies \Delta^\dagger \up
      % \delta\ \alpha_\mu \slash \J{TOP} \oplus \delta' \oplus \delta^\dagger \satisfies (e_a \J{,} e_r) \hastype \tau^\dagger 
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies \Delta \cup \Delta^\dagger \up
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies (e_a \J{,} e_r) \hastype \tau^\dagger 
      )
      )
      \implies
      e_r \cong e_f\J{(}e_a\J{)}
      )
    }
  \end{mathpar}
\end{requirement}

\noindent
The requirement of locale sequence duality has an ungodly amount of symbolic clutter,
but the essence of its meaning is actually quite simple and natural.
It says that a function inhabits all locales in some 
some collection if and only if for any input/ouput pair, if the pair inhabits some
locale from the dual collection, then the function applied to the input
gives the output. 
The requirement is essential to ensure that rewriting inferred locales
to construct least fixed point types does not violate soundness.
